ProbabilityBoundsAnalysis.jl:\\ Arithmetic with sets of distributions

### 2022-04-19 paper/paper.tex
removes subfig
adds further typeographic improvements

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{caption}
\usepackage{amsmath}
\usepackage{xcolor}
%\bibliographystyle{juliacon}
\usepackage[labelformat=simple]{subfig}
\renewcommand\thesubfigure{\arabic{subfigure}}
\renewcommand{\subfigurename}{Figure}
\begin{document}
\input{header}
\maketitle
\begin{abstract}
Probability bounds analysis combines interval arithmetic with probability theory, and provides a representation of sets of distributions in structures called probability boxes (p-boxes). P-boxes generalise both distribution functions and intervals, and return interval bounds on all probabilistic quantities, for example sample realisations, cdfs, and probability measures are all intervals. This framework also allows for the comprehensive propagation of probabilities through calculations in a rigorous way, in a similar fashion that interval arithmetic does for sets of real values. \texttt{ProbabilityBoundsAnalysis.jl} provides a rigorous arithmetic of random variables, where both marginal (univariate) distributions and dependency information can be known, partially known or missing entirely. We describe the main theoretical elements of probability bounds analysis, and provide a simplified implementation of the method in code snippets which can be readily evaluated in the Julia command terminal.
\end{abstract}
\section{Introduction}
\label{sec:intro}
An arithmetic of probability distributions has long held interest among mathematicians and scientists. Indeed it was Kolmogorov who originally asked about the result of an operation between two random variables without knowing their joint distribution. This was answered for the sum by Makarov~\cite{makarov1982estimates}, who showed the result was a set of distributions and was able to provide bounds on this function. Sklar, Schweizer and Frank generalised this result to other positive binary operations~\cite{frank1987best,schweizer2011probabilistic}. In this pursuit they created copulas, a general way to encode probabilistic dependence independently from marginals, which are now essential objects in probabilistic modelling. Williamson and Downs~\cite{williamson1990probabilistic} described an algorithm for efficiently performing these arithmetic operations, which can give guaranteed bounds on probability distributions in terms of an upper and lower cumulative distribution function (cdf). They called their method \textit{probabilistic arithmetic} and their sets of distributions \textit{dependency bounds}. Since then the method has been generalised~\cite{ferson2015constructing,ferson1996whereof,ferson2004arithmetic} to most of the base binary and unary operations that would be present in a programming language. Probability boxes (p-boxes) are the name now given to these structures.

@@

\end{figure*}
Imprecise probability is effectively a generalisation of probability theory where uncertainty can be expressed about the probability measure. This is particularly relevant when information is scarce, unreliable, vague, conflicting, or imprecise: cases where defining a unique probability distribution is difficult. P-boxes are one of many ways to describe a set of distributions. Others include: Dempster-Shafer structures~\cite{dempster2008upper,shafer1976mathematical}, random sets~\cite{molchanov2005theory}, possibility distributions~\cite{zadeh1978fuzzy,dubois1988possibility, hose2019possibilistic}, lower previsions~\cite{troffaes2014lower}, and credal sets~\cite{levi1983enterprise}. These structures were discovered independently, but are often synonymous (with different degrees of generality) and can be transformed from one to another. Imprecise probability unifies the underlying theories. For an overview of the theory, and a formal description of uncertainty and information in terms of these structures, work by George Klir~\cite{klir2013uncertainty} is recommended.
\subsection{Automatic uncertainty in a programming language: why Julia?}
A potential long term goal of probability bounds analysis is to create a uncertain programming language, where any computer variable may be represented as an interval, distribution, p-box, or other imprecise quantity. Such a framework would allow for uncertain extensions of deterministic functions to be computed in an automatic, rigorous, and tight fashion. Those with little knowledge of probability theory could add uncertainty to their computations with minimal effort. Stochastic computation often require specialist training to be used effectively, and are usually not rigorous and are famously computationally expensive. \texttt{ProbabilityBoundsAnalysis.jl} provides the arithmetic basis for such a framework. In this section we give some arguments why Julia~\cite{bezanson2017julia} is an ideal target language for such a framework.

@@

\subsubsection{Metaprogramming}
An uncertain programming language would require \textit{source-to-source} transformations at times. For example, one way to evaluate a programs control-flow (\texttt{if-else} conditions) with uncertain inputs is to use metaprogramming. The programs control flow could be unravelled, and with each possible conditional branch evaluated for the possible instance of the variable, and then combined as an output. For example consider the following function
\begin{lstlisting}[language = Julia]
  function unc_flow(x)
    if x < 0 
      return x - 1
    else 
      return x
    end
  end
\end{lstlisting}
\noindent An interval evaluation of $\text{unc\_flow}([-1,1])$ would be the union of the output of the different branches, the interval union $[-2, -1) \cup [0, 1]$. For a language like Python, this would require an uncertainty pre-compiler (like that proposed in~\cite{gray2019computing}) whereby the python code is first parsed using a separate program, and then evaluated in Python. In Julia, this can be handled natively with its metaprogramming capabilities.
\begin{figure*}[htp]

@@

\fi
\section{Probability Boxes}
\label{sec:pboxes}
A real-valued random variable is characterised by its distribution function $F$, which is a monotonically increasing function from the real numbers onto the interval $[0, 1]$ such that the value of the function at negative infinity is zero and the value of the function at positive infinity is one.  A p-box consists of a pair of such functions that are used to circumscribe an imprecisely known distribution function $F$. The p-box, in its simplest form consisting of the pair of bounding functions, identifies a set of probability distributions that lie entirely within these bounds. Additional information about the random variable may be available, such as bounds on its mean and variance and its distribution family, which may be used to further restrict the set of distributions. A p-box is thus defined by the following three constraints: (1) two bounding cumulative distribution functions (cdf), (2) interval bounds on the mean and variance, and (3) a collection of distribution families:
\begin{enumerate}
  \item $\begin{aligned}[t]
    \underline{F}(x) \leq F(x) \leq \overline{F}(x),
  \end{aligned}$
  \item $\begin{aligned}[t]
    \mu &\in [\underline{ \mu }, \overline{ \mu }]
    \sigma^2 &\in [\underline{\sigma}^2 , \overline{\sigma}^2],
  \end{aligned}$
  \item $\begin{aligned}[t]
      F \in \bm{F}.
  \end{aligned}$
\end{enumerate}
\noindent That is, a random variable is a member of a p-box if its cdf $F$ falls within the cdf bounds of the p-box $F(x) \in [\underline{F}(x), \overline{F}(x)]$ for all $x$, its moments are inside the interval moments of the p-box, and it belongs to a family of distribution functions (e.g. normal, uniform) considered by the p-box. Some of the constraints may be missing. For example, if the distribution family is unknown then the set is defined solely from the cdf and moment bounds. Such p-boxes are sometimes called non-parametric, since its members do not belong to any particular class of distribution. Some constraints may also be inferred from others. For example, the interval moments may be bounded from the cdf bounds, and cdf may be bounded from moment information (explored further in section \ref{section:Moments}). Figure \ref{fig:figure1} (scripts for figures found in \texttt{/examples/JuliaCon}) shows an example of a distribution function, interval and a p-box. The grey shaded region in the interval and p-box bounds an infinite collection of probability distributions.

Real-valued statistics and features of distributions typically become intervals for p-boxes. The cdf of a p-box is
\begin{equation*}
  [\underline{F}(x), \overline{F}(x)] .
\end{equation*}
\noindent A sample realisation of the p-box may be drawn using the inverses of the bounding cdfs
\begin{equation*}
  [\overline{F}^{-1}(\alpha), \underline{F}^{-1}(\alpha)]
\end{equation*}
\noindent where $\alpha \sim U(0,1)$ is a sample from a uniform distribution. The probability measure,  a function which returns the probability that the random variable is in some set, on some interval $U = [a, b]$ is bounded as follows
\begin{align*}
  \underline{\mathbb{P}}(U) &= \text{max}(0, \underline{F}(b) - \overline{F}(a)),\\ 
  \overline{\mathbb{P}}(U)  &= \overline{F}(b) - \underline{F}(a),
\end{align*}
\noindent where the max operator is required when $\underline{F}(b) < \overline{F}(a)$. Note the same can nearly be achieved by using the precise formula for the probability measure, $\mathbb{P}(U) = F(b)- F(a)$, and using interval arithmetic.
\begin{figure*}[htp]

@@

  \caption{Various parametrically constructed p-boxes with interval parameters.}
  \label{fig:figure3}
\end{figure*}
P-boxes generalise precise distributions and intervals in the following way. A distribution is a p-box with a precise cdf and moments, i.e. when
\begin{align*}
  \underline{F}(x) &= \overline{F}(x), \\ 
  \underline{\mu}  &= \overline{\mu}, \\ 
  \underline{\sigma}^2 &= \overline{\sigma}^2 .
\end{align*}
\noindent An interval $[a,b]$ is a p-box whose bounds are step functions
\begin{align*}
    \underline{F}(x) &= \epsilon_{b}(x) ,\\
    \overline{F}(x) &= \epsilon_{a}(x) ,
\end{align*}
\noindent where $\epsilon_k$ is
\begin{equation*}
   \epsilon_k(x) = \begin{cases} 0 &\text{when } x < k, \\ 1 &\text{when } x \geq k. \end{cases} 
\end{equation*}
\noindent Moreover, theoretical bounds on the mean and variance of an interval can be found~\cite{ferson2002ramas}
\begin{align*}
  \mu &\in [a, b], \\
  \sigma^2 &\in [0, (b - a)^{2}/4].
\end{align*}
\noindent That is, it is not possible to find a distribution inside $[a, b]$ with a variance greater than $(b-a)^{2}/4$. The lower bound on the variance is zero, since scalars are also included in the interval. Under this definition of an interval, a random sample will always be the interval $[a, b]$, the cdf returns 0 when $x < a$, the interval $[0,1]$ when $a \leq x < b$, and $1$ when $b \leq x$. Further, the probability measure of the interval $X = [a, b]$ is
\begin{equation*}
  \mathbb{P}_{X}(U) = \begin{cases}
    0 & \text{when } U \cap [a,b] = \emptyset, \\
    1 & \text{when } [a,b] \subseteq U,  \\
    [0, 1] & \text{otherwise.}\\
  \end{cases}
\end{equation*}
\noindent That is, if the set $U$ does not intersect the interval $X$, $\mathbb{P}_{X}=0$, i.e. $U$ certainly does not contain any of the random variables. If $X$ is fully contained in $U$, $\mathbb{P}_{X}=1$, i.e. $U$ certainly contains all of the random variables. And finally if $U$ intersects (but does not contain) $X$, the probability measure is the vacuous probability interval $\mathbb{P}_{X} = [0, 1]$, i.e. we are completely uncertain about the containment. Note that the $P_{X} = \{0, 1, [0,1]\}$ is due to the interval bounds being step functions. Generally p-boxes may yield other probability intervals.
\subsection{Outer representations of p-boxes}
\label{sec:outer_approx}
An important feature of probability bounds analysis is how distribution functions and p-boxes are represented. Generally, analytical solutions for cdfs are not readily available, for example the normal distribution's cdf can only be found by integrating the probability density, usually with quadrature. However, even if the functions were available analytically, when the variables are used in an arithmetic operation the output distribution will not necessarily belong to the same family. Therefore, we require a representation of these continuous functions which is not dependent on the distribution family (for example as used in polynomial chaos expansion), and ideally is robust in the sense that it can produce an interval error for the numerical representation. Such a representation of a probability is sometimes called an "outer approximation", in the same fashion an interval is an outer approximation of a floating point number. Such outer approximations contract to the exact value or function as more computational resources are used.

We follow the representation first introduced by Williamson and Downs~\cite{williamson1990probabilistic}, where a discrete upper and lower approximations of distribution functions are constructed using the inverse cdfs. Note that when considering inverse cdfs $\underline{F}^{-1}(p) \geq F^{-1}(p) \geq \overline{F}^{-1}(p)$, for $p \in [0,1]$. Beginning vector indexing by $1$, an outer approximation using two finite vectors $u$ and $d$ of length $N$ is constructed by evaluating the inverse cdfs for uniformly spaced probability levels $p_{i} = \frac{i-1}{N}$ for $i = 1, ... , N+1$, e.g. $p_{i} = [0, 0.25, 0.5, 0.75, 1]$ for $N = 4$. The vectors $u$ and $d$ are defined as follows
\begin{align*}
  u[i] &= \overline{F}^{-1}(p_{i}), \\ 
  d[i] &= \underline{F}^{-1}(p_{i+1}), & \text{for } i = 1, ..., N .
\end{align*}
The vectors $u$ and $d$ are a collection of "real" or "physical" evaluations of a the random variable $X$, as opposed to cdf evaluations. Although the probability levels $p$ are uniformly spaced, $u$ and $d$ are generally not. The vectors $u$ and $d$ can then serve as bounds on the monotonic functions $\overline{F}^{-1}$ and $\underline{F}^{-1}$
\begin{align*}
  u[i] &\leq \overline{F}^{-1}(p), & p_{i} \in [p_{i}, p_{i+1} ),\\ 
  d[i] &\geq \underline{F}^{-1}(p), & p_{i} \in (p_{i}, p_{i+1}].
\end{align*}
That is, in the region $[p_{i}, p_{i+1})$, $u[i]$ is a lower bound on $\overline{F}^{-1}$ and $d[i]$ is an upper bound on $\underline{F}^{-1}$ in $(p_{i}, p_{i+1}]$. In the case that the p-box characterises a precise distribution and $\overline{F}^{-1}(p) = \underline{F}^{-1}(p)$ gives:  $u[i+1] = d[i]$. Figure \ref{fig:figure2} shows examples of a distribution and a p-box with various outer representations.

@@

  struct pbox{T <: Real}
    u :: Vector{T}    # left discrete inverse
    d :: Vector{T}    # right discrete inverse
    m :: Interval{T}  # interval mean
    v :: Interval{T}  # interval variance
    shape :: String   # shape
  end
\end{lstlisting}
\noindent where the vectors $u$ and $d$ are the "up" and "down", or the "left" and "right", edges of the p-box. There can be confusion since $\overline{F}$ is the upper bound on the cdf, but $\overline{F}^{-1}$ is the lower bound on the random variable, i.e. the left edge of the p-box. We therefore call $\overline{F}$ the "up" or "left" edge, and $\underline{F}$ the "down" or the "right" edge. The intervals $m$ and $v$ are the interval moments, and the $shape$ is a string containing any distribution family information. An example of a normal distribution constructor is as follows
\begin{lstlisting}[language = Julia]
  using Distributions
  const steps = 200

  function normal(m :: Interval, std :: Interval)

    ps = range(0, 1, length = steps+1)

@@

    return pbox(u, d, m, std^2, "normal")
  end
\end{lstlisting}
Other distributional p-boxes are constructed in a similar way. An evaluation of the cdf in terms of this representation is as follows
\begin{figure*}[htp]
  \centering
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig6/fig6_pbox1.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig6/fig6_pbox2.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig6/fig6_pbox3.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig6/fig6_pbox4.pdf}
  \caption{Examples of distribution-free p-boxes, constructed from various range and moment constrains.}
  \label{fig:figure6}
\end{figure*}
\begin{lstlisting}[language = Julia]
  function cdf(F :: pbox, x :: Real)

@@

    return interval(cdf_d, cdf_u)
  end
\end{lstlisting}
A horizontal cut (an evaluation of the inverse cdf) and a sample of a p-box is
\begin{lstlisting}[language = Julia]
  import Base: rand

@@

  # a random sample is a random cut
  rand(F :: pbox, N :: Int) = cut.([F], rand(N))
\end{lstlisting}
And finally the probability measure, which we call the \textit{mass} for short
\begin{lstlisting}[language = Julia]
  function mass(F :: pbox, x :: Interval)

@@

Covariance, and subsequently the pearson correlation coefficient, is insufficient to fully determine the dependence between two random variables. Like for the univariate moments, a single multivariate distribution cannot be prescribed for a given value of the covariance without making distributional assumptions (like normality). A more descriptive (and general) model for dependence is required to exactly specify a multivariate distribution.

Copulas~\cite{nelsen2007introduction} provide a solution to this problem. A copula is a multivariate cdf with uniform marginals on the interval $[0, 1]$. A copula is a multivariate distribution which has all marginal (univariate) information stripped away, leaving only the dependency. All dependence information can therefore be encoded in a copula exactly and completely separately from its marginals. Often multivariate problems in probability can be reduced to an analysis of copulas.

>>>>

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{caption}
\usepackage{amsmath}
\usepackage{xcolor}
%\bibliographystyle{juliacon}
\begin{document}
\input{header}
\maketitle
\begin{abstract}
Probability bounds analysis combines interval arithmetic with probability theory, and provides a representation of sets of distributions in structures called probability boxes (p-boxes). P-boxes generalise both distribution functions and intervals, and return interval bounds on all probabilistic quantities, for example sample realisations, cdfs, and probability measures are all intervals. This framework also allows for the comprehensive propagation of probabilities through calculations in a rigorous way, in a similar fashion that interval arithmetic does for sets of real values. \texttt{ProbabilityBoundsAnalysis.jl} provides a rigorous arithmetic of random variables, where both marginal (univariate) distributions and dependency information can be known, partially known or missing entirely. We describe the main theoretical elements of probability bounds analysis, and provide a simplified implementation of the method in code snippets which can be readily evaluated in the Julia command terminal.
\end{abstract}
\section{Introduction}
\label{sec:intro}
An arithmetic of probability distributions has long held interest among mathematicians and scientists. Indeed it was Kolmogorov who originally asked about the result of an operation between two random variables without knowing their joint distribution. This was answered for the sum by Makarov~\cite{makarov1982estimates}, who showed the result was a set of distributions and was able to provide bounds on this function. Sklar, Schweizer and Frank generalised this result to other positive binary operations~\cite{frank1987best,schweizer2011probabilistic}. In this pursuit they created copulas, a general way to encode probabilistic dependence independently from marginals, which are now essential objects in probabilistic modelling. Williamson and Downs~\cite{williamson1990probabilistic} described an algorithm for efficiently performing these arithmetic operations, which can give guaranteed bounds on probability distributions in terms of an upper and lower cumulative distribution function (cdf). They called their method \textit{probabilistic arithmetic} and their sets of distributions \textit{dependency bounds}. Since then the method has been generalised~\cite{ferson2015constructing,ferson1996whereof,ferson2004arithmetic} to most of the base binary and unary operations that would be present in a programming language. Probability boxes (p-boxes) are the name now given to these structures.

@@

\end{figure*}
Imprecise probability is effectively a generalisation of probability theory where uncertainty can be expressed about the probability measure. This is particularly relevant when information is scarce, unreliable, vague, conflicting, or imprecise: cases where defining a unique probability distribution is difficult. P-boxes are one of many ways to describe a set of distributions. Others include: Dempster-Shafer structures~\cite{dempster2008upper,shafer1976mathematical}, random sets~\cite{molchanov2005theory}, possibility distributions~\cite{dubois1988possibility, hose2019possibilistic}, lower previsions~\cite{troffaes2014lower}, and credal sets~\cite{levi1983enterprise}. These structures were discovered independently, but are often synonymous (with different degrees of generality) and can be transformed from one to another. Imprecise probability unifies the underlying theories. For an overview of the theory, and a formal description of uncertainty and information in terms of these structures, work by George Klir~\cite{klir2013uncertainty} is recommended.
\subsection{Automatic uncertainty in a programming language: why Julia?}
A potential long term goal of probability bounds analysis is to create a uncertain programming language, whereby any computer variable may be represented as an interval, distribution, p-box, or other imprecise quantity. Such a framework would allow for uncertain extensions of deterministic functions to be computed in an automatic, rigorous, and tight fashion. Those with little knowledge of probability theory could add uncertainty to their computations with minimal effort. Stochastic computation often require specialist training to be used effectively, and are usually not rigorous and are famously computationally expensive. \texttt{ProbabilityBoundsAnalysis.jl} provides the arithmetic basis for such a framework. In this section we give some arguments why Julia~\cite{bezanson2017julia} is an ideal target language for such a framework.

@@

\subsubsection{Metaprogramming}
An uncertain programming language could require \textit{source-to-source} transformations at times. For example, one way to evaluate a programs control-flow (\texttt{if-else} conditions) with uncertain inputs is to use metaprogramming. The programs control flow could be unravelled, and with each possible conditional branch evaluated for the possible instance of the variable, and then combined as an output. For example consider the following function
\begin{lstlisting}[language = Julia]
  function unc_flow(x)
    if x < 0 
      return x - 1
    else 
      return x
    end
  end
\end{lstlisting}
\noindent An interval evaluation of $\text{unc\_flow}([-1,1])$ would be the union of the output of the different branches, the interval union $[-2, -1) \cup [0, 1]$. For a language like Python, this would require an uncertainty pre-compiler (like that proposed in~\cite{gray2019computing}) whereby the python code is first parsed using a separate program, and then evaluated in Python. In Julia, this can be handled natively with its metaprogramming capabilities.
\begin{figure*}[htp]

@@

\fi
\section{Probability Boxes}
\label{sec:pboxes}
A real-valued random variable is characterised by its distribution function $F$, which is a monotonically increasing function from the real numbers onto the interval $[0, 1]$ such that the value of the function at negative infinity is zero and the value of the function at positive infinity is one.  A p-box consists of a pair of such functions that are used to circumscribe an imprecisely known distribution function $F$. The p-box, in its simplest form consisting of the pair of bounding functions, identifies a set of probability distributions that lie entirely within these bounds. Additional information about the random variable may be available, such as bounds on its mean and variance and its distribution family, which may be used to further restrict the set of distributions. A p-box is thus defined by the following three constraints: (1) two bounding cumulative distribution functions (cdf), (2) interval bounds on the mean and variance, and (3) a collection of distribution families:
\begin{enumerate}
  \item $\begin{aligned}[t]
    \underline{F}(x) \leq F(x) \leq \overline{F}(x),
  \end{aligned}$
  \item $\begin{aligned}[t]
    \mu \in [\underline{ \mu }, \overline{ \mu }], \;\;\;
    \sigma^2\in [\underline{\sigma}^2 , \overline{\sigma}^2],
  \end{aligned}$
  \item $\begin{aligned}[t]
      F \in \bm{F}.
  \end{aligned}$
\end{enumerate}
\noindent That is, a random variable is a member of a p-box if its cdf $F$ falls within the cdf bounds of the p-box $F(x) \in [\underline{F}(x), \overline{F}(x)]$ for all $x$, its moments are inside the interval moments of the p-box, and it belongs to the same distribution family (e.g. normal, uniform) as the p-box. Some of the constraints may be missing. For example, if the distribution family is unknown then the set is defined solely from the cdf and moment bounds. Such p-boxes are sometimes called non-parametric, since its members do not belong to any particular class of distribution. Some constraints may also be inferred from others. For example, the interval moments may be bounded from the cdf bounds, and cdf may be bounded from moment information (explored further in section \ref{section:Moments}). Figure \ref{fig:figure1} (scripts for figures found in \texttt{/examples/JuliaCon}) shows an example of a distribution function, interval and a p-box. The grey shaded region in the interval and p-box bounds an infinite collection of probability distributions.

Real-valued statistics and features of distributions typically become intervals for p-boxes. The cdf of a p-box is
\begin{equation}
  \label{eq:cdf}
  [\underline{F}(x), \overline{F}(x)] .
\end{equation}
\noindent A sample realisation of the p-box may be drawn using the inverses of the bounding cdfs
\begin{equation}
  \label{eq:sample}
  [\overline{F}^{-1}(\alpha), \underline{F}^{-1}(\alpha)]
\end{equation}
\noindent where $\alpha \sim U(0,1)$ is a sample from a uniform distribution. The probability measure,  a function which returns the probability that the random variable is in some set, on some interval $U = [a, b]$ is bounded as follows
\begin{align}
  \label{eq:measure}
  \underline{\mathbb{P}}(U) &= \text{max}(0, \underline{F}(b) - \overline{F}(a)), \notag\\ 
  \overline{\mathbb{P}}(U)  &= \overline{F}(b) - \underline{F}(a),
\end{align}
\noindent where the max operator is required when $\underline{F}(b) < \overline{F}(a)$. Note the same can nearly be achieved by using the precise formula for the probability measure, $\mathbb{P}(U) = F(b)- F(a)$, and using interval arithmetic.
\begin{figure*}[htp]

@@

  \caption{Various parametrically constructed p-boxes with interval parameters.}
  \label{fig:figure3}
\end{figure*}
P-boxes generalise precise distributions and intervals in the following way. A distribution is a p-box with a precise cdf and moments, i.e. when
\begin{align*}
  \underline{F}(x) &= \overline{F}(x), \\ 
  \underline{\mu}  &= \overline{\mu}, \\ 
  \underline{\sigma}^2 &= \overline{\sigma}^2 .
\end{align*}
\noindent An interval $[a,b]$ is a p-box whose bounds are step functions
\begin{align*}
    \underline{F}(x) &= \epsilon_{b}(x) ,\\
    \overline{F}(x) &= \epsilon_{a}(x) ,
\end{align*}
\noindent where $\epsilon_k$ is
\begin{equation*}
   \epsilon_k(x) = \begin{cases} 0 &\text{when } x < k, \\ 1 &\text{when } x \geq k. \end{cases} 
\end{equation*}
\noindent Moreover, theoretical bounds on the mean and variance of an interval can be found~\cite{ferson2002ramas}
\begin{align*}
  \mu &\in [a, b], \\
  \sigma^2 &\in [0, (b - a)^{2}/4].
\end{align*}
\noindent That is, it is not possible to find a distribution inside $[a, b]$ with a variance greater than $(b-a)^{2}/4$. The lower bound on the variance is zero, since scalars are also included in the interval. Under this definition of an interval, a random sample will always be the interval $[a, b]$, the cdf returns 0 when $x < a$, the interval $[0,1]$ when $a \leq x < b$, and $1$ when $b \leq x$. Further, the probability measure of the interval $X = [a, b]$ is
\begin{equation*}
  \mathbb{P}_{X}(U) = \begin{cases}
    0 & \text{when } U \cap [a,b] = \emptyset, \\
    1 & \text{when } [a,b] \subseteq U,  \\
    [0, 1] & \text{otherwise.}\\
  \end{cases}
\end{equation*}
\noindent That is, if the set $U$ does not intersect the interval $X$, $\mathbb{P}_{X}=0$, i.e. $U$ certainly does not contain any of the random variables. If $X$ is fully contained in $U$, $\mathbb{P}_{X}=1$, i.e. $U$ certainly contains all of the random variables. And finally if $U$ intersects (but does not contain) $X$, the probability measure is the vacuous probability interval $\mathbb{P}_{X} = [0, 1]$, i.e. we are completely uncertain about the containment. Note that the $P_{X} = \{0, 1, [0,1]\}$ is due to the interval bounds being step functions. Generally p-boxes may yield other probability intervals.
\subsection{Outer representations of p-boxes}
\label{sec:outer_approx}
An important feature of probability bounds analysis is how distribution functions and p-boxes are represented. Generally, analytical solutions for cdfs are not readily available, for example the normal distribution's cdf can only be found by integrating the probability density, usually with quadrature. However, even if the functions were available analytically, when the variables are used in an arithmetic operation the output distribution will not necessarily belong to the same family. Therefore, we require a representation of these continuous functions which is not dependent on the distribution family, and ideally is robust in the sense that it can produce an interval error for the numerical representation. Such a representation of a probability is sometimes called an ``outer approximation'', in the same fashion an interval is an outer approximation of a floating point number. Such outer approximations contract to the exact value or function as more computational resources are used.

We follow the representation first introduced by Williamson and Downs~\cite{williamson1990probabilistic}, where a discrete upper and lower approximations of distribution functions are constructed using the inverse cdfs. Note that when considering inverse cdfs $\underline{F}^{-1}(p) \geq F^{-1}(p) \geq \overline{F}^{-1}(p)$, for $p \in [0,1]$. Beginning vector indexing by $1$, an outer approximation using two finite vectors $u$ and $d$ of length $N$ is constructed by evaluating the inverse cdfs for uniformly spaced probability levels $p_{i} = \frac{i-1}{N}$ for $i = 1, ...\, , N+1$, e.g. $p_{i} = [0, 0.25, 0.5, 0.75, 1]$ for $N = 4$. The vectors $u$ and $d$ are defined as follows
\begin{align*}
  u[i] &= \overline{F}^{-1}(p_{i}), \\ 
  d[i] &= \underline{F}^{-1}(p_{i+1}), & \text{for } i = 1, ... \,, N .
\end{align*}
The vectors $u$ and $d$ are a collection of ``real'' or ``physical'' evaluations of a the random variable $X$, as opposed to cdf evaluations. Although the probability levels $p$ are uniformly spaced, $u$ and $d$ are generally not. The vectors $u$ and $d$ can then serve as bounds on the monotonic functions $\overline{F}^{-1}$ and $\underline{F}^{-1}$
\begin{align*}
  u[i] &\leq \overline{F}^{-1}(p), & p_{i} \in [p_{i}, p_{i+1} ),\\ 
  d[i] &\geq \underline{F}^{-1}(p), & p_{i} \in (p_{i}, p_{i+1}].
\end{align*}
That is, in the region $[p_{i}, p_{i+1})$, $u[i]$ is a lower bound on $\overline{F}^{-1}$ and $d[i]$ is an upper bound on $\underline{F}^{-1}$ in $(p_{i}, p_{i+1}]$. In the case that the p-box characterises a precise distribution and $\overline{F}^{-1}(p) = \underline{F}^{-1}(p)$ gives:  $u[i+1] = d[i]$. Figure \ref{fig:figure2} shows examples of a distribution and a p-box with various outer representations.

@@

  struct pbox{T <: Real}
    u :: Vector{T}    # left discrete inverse
    d :: Vector{T}    # right discrete inverse
    m :: Interval{T}  # interval mean
    v :: Interval{T}  # interval variance
    shape :: String   # shape
  end
\end{lstlisting}
\noindent where the vectors $u$ and $d$ are the ``up'' and ``down'', or the ``left'' and ``right'', edges of the p-box. There can be confusion since $\overline{F}$ is the upper bound on the cdf, but $\overline{F}^{-1}$ is the lower bound on the random variable, i.e. the left edge of the p-box. We therefore call $\overline{F}$ the ``up'' or ``left'' edge, and $\underline{F}$ the ``down'' or the ``right'' edge. The intervals $m$ and $v$ are the interval mean and variance, and $shape$ is a string containing any distribution family information. An example of a normal distribution constructor is as follows
\begin{lstlisting}[language = Julia]
  using Distributions
  const steps = 200

  function normal(m :: Interval, std :: Interval)
    ps = range(0, 1, length = steps+1)

@@

    return pbox(u, d, m, std^2, "normal")
  end
\end{lstlisting}
Other distributional p-boxes are constructed in a similar way. An evaluation of the cdf (Equation \ref{eq:cdf}) in terms of this representation is as follows
\begin{figure*}[htp]
  \centering
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig6/fig6_pbox1.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig6/fig6_pbox2.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig6/fig6_pbox3.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig6/fig6_pbox4.pdf}
  \caption{Examples of distribution-free p-boxes, constructed from various range and moment constrains.}
  \label{fig:figure6}
\end{figure*}
\begin{lstlisting}[language = Julia]
  function cdf(F :: pbox, x :: Real)

@@

    return interval(cdf_d, cdf_u)
  end
\end{lstlisting}
A horizontal cut (an evaluation of the inverse cdf) and a sample (Equation \ref{eq:sample}) of a p-box is
\begin{lstlisting}[language = Julia]
  import Base: rand

@@

  # a random sample is a random cut
  rand(F :: pbox, N :: Int) = cut.(F, rand(N))
\end{lstlisting}
And finally the probability measure (Equation \ref{eq:measure}), which we call the \textit{mass} for short
\begin{lstlisting}[language = Julia]
  function mass(F :: pbox, x :: Interval)

@@

Covariance, or Pearson's correlation coefficient, is insufficient to fully determine the dependence between two random variables. Like for the univariate moments, a single multivariate distribution cannot be prescribed for a given value of the covariance without making distributional assumptions (like normality). A more descriptive (and general) model for dependence is required to exactly specify a multivariate distribution.

Copulas~\cite{nelsen2007introduction} provide a solution to this problem. A copula is a multivariate cdf with uniform marginals on the interval $[0, 1]$. A copula is a multivariate distribution which has all marginal (univariate) information stripped away, leaving only the dependency. All dependence information can therefore be encoded in a copula exactly and completely separately from its marginals. Often multivariate problems in probability can be reduced to an analysis of copulas. A bivariate copula (2-copula) $C$ is any function $C:[0,1]^2 \rightarrow [0,  1]$ with the following properties
\begin{enumerate}
  \item Grounded: $C(0,v) = C(u,0) = 0$,
  \item Uniform margins: $C(u,1) = u; \;C(1,v) = v$,
  \item 2-increasing: \\ $C(u_{2},v_{2}) - C_{}(u_{2}, v_{1}) - C(u_{1}, v_{2}) + C(u_{1}, v_{1}) \ge 0$\\ for all $0 \le u_{1} \le u_{2} \le 1$ and $0 \le v_{1} \le v_{2} \le 1$ .
\end{enumerate}
\noindent Three important 2-copulas are
\begin{align*}
  W(u,v) &= \mathrm{max}( u + v-1,0),  \\
  \Pi(u,v) &= uv, \\
  M(u,v) &= \mathrm{min}(u,v),
\end{align*}
\noindent where $W$ encodes maximally negative correlation, $\Pi$ encodes stochastic independence, and $M$ encodes maximally positive correlation. Moreover, $W$ and $M$ are bounds on all 2-copulas: $W(u,v) \leq C(u,v) \leq M(u,v)$. A bivariate cdf $F_{XY}$ can be constructed from two marginal cdfs $F_{X}$ and $F_{Y}$ and copula $C$ using Sklar's theorem
\begin{equation*}
  F_{XY}(x,y) = C(F_{X}(x), F_{Y}(y)),
\end{equation*}
\noindent where the constructed bivariate distribution will follow the dependence encoded in the copula and have the prescribed marginals. Sklar's theorem also works in reverse, where the dependence can be extracted from a multivariate distribution: 
\begin{equation*}
  C(u,v) = F_{XY}(F^{-1}_{X}(u), F^{-1}_{Y}(v)).
\end{equation*}
\noindent Copulas also give a very convenient way to sample from $F_{XY}$ in a generalisation of inverse transform sampling, where first samples are drawn from the copula $(u,v) \sim C$, and then transformed through the inverses of the marginals $(F^{-1}_{X}(u), F^{-1}_{Y}(v))$.
\begin{figure*}[t!]
  \centering
  \includegraphics[width=0.45\textwidth]{../examples/JuliaCon/fig7/biv_cdf.pdf}\hfill
  \includegraphics[width=0.45\textwidth]{../examples/JuliaCon/fig7/samples.pdf}
  \caption{Cdf bounds of a bivariate p-box (right) with $X \sim \text{Beta}([4,8], 3)$, $Y \sim \text{N}([0,2], 2)$, and $\rho_{XY} = -0.8$, and 100 samples (left).}
  \label{fig:figure7}
\end{figure*}
